Miniproject - Analytic Number Theory

### 2021-12-19 Algebraic geometry/submission.tex
Added introduction & conclusion to ANT

    Then for an ideal $I \leq R$ have that the list of all monomials $x_0^{\alpha_0}, ..., x_n^{\alpha_n} \notin \mathrm{lt}(I)$ not contained in $\mathrm{lt}(I)$ are a $k$-vector space basis of $R/I$.

    Here $\mathrm{lt}(I)$ is the leading term ideal of $I$, i.e. the ideal generated by the leading terms of all $f \in I$, w.r.t. $\preceq$.
\end{proposition}
\begin{proof}
    See \cite{kreuzer}.
\end{proof}
To apply this, first of all we have to collect information about the leading term ideal of $\I(\Gr(2, n))$.
This is done in the following lemma.
\begin{lemma}
    \label{prop:leading_term_ideal_grassmanian}
    Define the graded reverse monomial ordering $\preceq$ on $R := k[x_{i, j} \ | \ i < j]$ where the variables $x_{i, j}$ are ordered lexicographically w.r.t. $(i, j)$, i.e.
    \begin{equation*}
        x_{i, j} \leq x_{u, v} \ :\Leftrightarrow \ (i, j) \leq_{\text{lex}} (u, v)
    \end{equation*}
    Moreover, let
    \begin{equation*}
        I := \langle x_{i, j}x_{u, v} + x_{i, v}x_{j, u} - x_{i, u}x_{j, v} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
    be the ideal defining $\phi(\Gr(d, V))$ that was considered above. Then
    \begin{equation*}
        \mathrm{lt}(I) = J := \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
\end{lemma}
\begin{proof}
    We show that the set $G = \{ x_{iv} x_{ju} - x_{iu} x_{jv} + x_{ij} x_{uv} \ | \ i < j < u < v \}$ is a Gröbner basis of $I$
    \footnote{In this proof, we write $x_{ij} := x_{i, j}$ to somewhat simplify notation.}.
    The claim then follows by the properties of Gröbner basis.

@@

    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we see that
    \begin{equation*}
        \mathrm{lt}(f)^d = \mathrm{lt}(f^d) \in J = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle
    \end{equation*}
    Hence
    \begin{equation*}
        \mathrm{lt}(f)^d \equiv x_{i, v} x_{j, u}
    \end{equation*}
    for some $i < j < u < v$.
    Clearly $x_{i, v} \perp x_{j, u}$, and it follows that
    \begin{equation*}
        x_{i, v} x_{j, u} \divides \mathrm{lt}(f)
    \end{equation*}
    Now consider
    \begin{equation*}
        f' := f - \frac {\mathrm{lt}(f)} {x_{i, v} x_{j, u}} (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v}) \in R
    \end{equation*}
    Clearly $\mathrm{lt}(f') \prec \mathrm{lt}(f)$ as $x_{i, v} x_{j, v}, x_{i, j} x_{u, v} \prec x_{i, v} x_{j, u} \equiv \mathrm{lt}(f)$.
    Furthermore
    \begin{equation*}
        (f')^d = \sum_{l = 0}^d {d \choose i} f^i (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v})^{d - i} \in I
    \end{equation*}
    so by induction hypothesis, find that $f' \in I$.
    It follows that also
    \begin{equation*}
        f = f'  - \frac {\mathrm{lt}(f)} {x_{i, v} x_{j, u}} (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v}) \in I
    \end{equation*}
    and the claim is shown.
\end{proof}
Now we have to study the structure of monomials generating $J = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle$.
At first glance, this seemed to me to be much simpler than the above, but in fact it turns out not to be, as there is some subtle combinatorial structure involved.
Hence, we will first introduce the combinatorial framework in which we will work now.
\begin{definition}
    Consider the graph $G_n = (V_n, E_n)$ where $V_n = \{ (i, j) \ | \ 0 \leq i < j \leq n \}$ and
    \begin{equation*}
        E_n = \{ \{ (i, v), (j, u) \} \ | \ 0 \leq i < j < u < v \leq n \}
    \end{equation*}
    Let further
    \begin{equation*}
        s_n(d) := |\{ I \subseteq V_n \ | \ I \ \text{independent set of size $d$}\}|
    \end{equation*}
\end{definition}
It might be possible to see from the definition how independent sets in this graph are connected to monomials in $J$.
If not, the connection is explained in the proof of the next theorem.
\begin{proposition}
    Let $d$ be the largest integer such that $s_n(d) \neq 0$, i.e. the size of the largest independent set in $G_n$.
    Then
    \begin{equation*}
        \dim(\Gr(2, n)) = d - 1
    \end{equation*}
    and
    \begin{equation*}
        \deg(\Gr(2, n)) = s_n(d)
    \end{equation*}
\end{proposition}
\begin{proof}
    The idea is to apply Macaulay's basis theorem to get information about the Hilbert function of $\Gr(2, n)$.
    To this end, observe that all the monomials generating $\mathrm{lt}(\I(\Gr(2, n)))$ are square-free.
    Hence, we can partition all monomials not in $\mathrm{lt}(\I(\Gr(2, n)))$ into sets depending on which variables occur in them.
    Doing this shows that leading coefficient of the Hilbert polynomial depends only on the groups with the maximal count of monomials.
    These monomials then form a maximal independent set in $G_n$, which allows us to relate degree and dimension of $\Gr(2, n)$ to $s_n(d)$.

>>>>

    Then for an ideal $I \leq R$ have that the list of all monomials $x_0^{\alpha_0}, ..., x_n^{\alpha_n} \notin \mathrm{lt}(I)$ not contained in the leading term ideal $\mathrm{lt}(I)$ are a $k$-vector space basis of $R/I$.

    Here $\mathrm{lt}(I)$ is the leading term ideal of $I$, i.e. the ideal generated by the leading terms of all $f \in I$, w.r.t. $\preceq$.
\end{proposition}
\begin{proof}
    See \cite{kreuzer}.
\end{proof}
To apply this, first of all we have to collect information about the leading term ideal of $I$.
This is done in the following lemma.
\begin{lemma}
    \label{prop:leading_term_ideal_grassmanian}
    Define the graded reverse monomial ordering $\preceq$ on $R := k[x_{i, j} \ | \ i < j]$ where the variables $x_{i, j}$ are ordered lexicographically w.r.t. $(i, j)$, i.e.
    \begin{equation*}
        x_{i, j} \leq x_{u, v} \ :\Leftrightarrow \ (i, j) \leq_{\text{lex}} (u, v)
    \end{equation*}
    Moreover, let
    \begin{equation*}
        I := \langle x_{i, j}x_{u, v} + x_{i, v}x_{j, u} - x_{i, u}x_{j, v} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
    be the ideal defining $\phi(\Gr(d, V))$ that was considered above. Then
    \begin{equation*}
        \mathrm{lt}(I) = J := \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
\end{lemma}
\begin{proof}
    We show that the set $G = \{ x_{iv} x_{ju} - x_{iu} x_{jv} + x_{ij} x_{uv} \ | \ i < j < u < v \}$ is a Gröbner basis of $I$
    \footnote{In this proof, we write $x_{ij} := x_{i, j}$ to somewhat simplify notation.}.
    The claim then follows by the properties of Gröbner basis.

@@

    By Lemma~\ref{prop:leading_term_ideal_grassmanian} we see that
    \begin{equation*}
        \mathrm{lt}(f)^d = \mathrm{lt}(f^d) \in J = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle
    \end{equation*}
    Hence
    \begin{equation*}
        \mathrm{lt}(f)^d \equiv t x_{i, v} x_{j, u}
    \end{equation*}
    for some $i < j < u < v$ and any monomial $t \in R$.
    Clearly $x_{i, v} \perp x_{j, u}$, and it follows that
    \begin{equation*}
        x_{i, v} x_{j, u} \divides \mathrm{lt}(f)
    \end{equation*}
    Now consider
    \begin{equation*}
        f' := f - \frac {\mathrm{lt}(f)} {x_{i, v} x_{j, u}} (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v}) \in R
    \end{equation*}
    Clearly $\mathrm{lt}(f') \prec \mathrm{lt}(f)$ as $x_{i, v} x_{j, v}, x_{i, j} x_{u, v} \prec x_{i, v} x_{j, u} \equiv \mathrm{lt}(f)$.
    Furthermore
    \begin{equation*}
        (f')^d = \sum_{l = 0}^d {d \choose i} f^i (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v})^{d - i} \in I
    \end{equation*}
    so by induction hypothesis, find that $f' \in I$.
    It follows that also
    \begin{equation*}
        f = f'  - \frac {\mathrm{lt}(f)} {x_{i, v} x_{j, u}} (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v}) \in I
    \end{equation*}
    and the claim is shown.
\end{proof}
Now we have to study the structure of monomials generating $J = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle$.
At first glance, this seemed to me to be much simpler than the above, but in fact it turns out not to be, as there is some subtle combinatorial structure involved.
Hence, we will first introduce the combinatorial framework in which we will work now.
\begin{definition}
    Consider the graph $G_n = (V_n, E_n)$ where $V_n = \{ (i, j) \ | \ 0 \leq i < j \leq n \}$ and
    \begin{equation*}
        E_n = \{ \{ (i, v), (j, u) \} \ | \ 0 \leq i < j < u < v \leq n \}
    \end{equation*}
    Let further
    \begin{equation*}
        s_n(d) := |\{ I \subseteq V_n \ | \ I \ \text{independent set of size $d$}\}|
    \end{equation*}
\end{definition}
It might be possible to see from the definition how independent sets in this graph are connected to monomials in $J$.
If not, the connection is explained in the proof of the next theorem.
\begin{proposition}
    Let $d$ be the largest integer such that $s_n(d) \neq 0$, i.e. the size of the largest independent set in $G_n$.
    Then
    \begin{equation*}
        \dim(\Gr(2, n)) = d - 1
    \end{equation*}
    and
    \begin{equation*}
        \deg(\Gr(2, n)) = s_n(d)
    \end{equation*}
\end{proposition}
\begin{proof}
    The idea is to apply Macaulay's basis theorem to get information about the Hilbert function of $\Gr(2, n)$.
    To this end, observe that all the monomials generating $\mathrm{lt}(\I(\Gr(2, n)))$ are square-free.
    Hence, we can partition all monomials not in $\mathrm{lt}(\I(\Gr(2, n)))$ into sets depending on which variables occur in them.
    Doing this shows that leading coefficient of the Hilbert polynomial depends only on the groups with the maximal count of monomials.
    These monomials then form a maximal independent set in $G_n$, which allows us to relate degree and dimension of $\Gr(2, n)$ to $s_n(d)$.

### 2021-12-19 Analytic number theory/submission.tex
Added introduction & conclusion to ANT

\title{Miniproject - Analytic Number Theory}
\author{Simon Pohmann}
\date{}

@@

\begin{document}
\maketitle
We use the convention that $\N = \{ n \in \Z \ | \ n \geq 0 \}$.
Further, we write $a \divides b$ if $a$ divides $b$ and $a \perp b$ if $a$ and $b$ are coprime.
Finally, let $\primes$ be the set of prime numbers in $\N$.

\section{Part I}

For convenience, we include the definition of a Dirichlet character from the task description first.
\begin{definition}
    Let $q \geq 2$, then a \emph{Dirichlet character (modulo $q$)} is a function $\chi: \N \to \C$ such that
    \begin{itemize}
        \item $\chi$ is completely multiplicative, so $\chi(a)\chi(b) = \chi(ab)$
        \item $\chi$ is periodic modulo $q$, so $\chi(n + q) = \chi(n)$
        \item $\chi(n) \neq 0$ if and only if $n \perp q$
    \end{itemize}
\end{definition}
First, we will give another characterization of Dirichlet characters.
\begin{lemma}[Characterization of Dirichlet characters]
    \label{prop:characterization_dirichlet_character}
    We have a one-to-one correspondence between Dirichlet characters modulo $q$ and group homomorphisms $(\Z/q\Z)^\units \to \C^\units$ via
    \begin{align*}
        \{ f: (\Z/q\Z)^\units \to \C^\units \ | \ f \ \text{group hom} \} &\to \{ \chi: \N \to \C \ | \ \chi \ \text{Dirichlet character mod $q$} \} \\
        f &\mapsto \chi_f := \left( \N \to \C, \ n \mapsto \begin{cases}
            f([n]) & \text{if $n \perp q$} \\
            0 & \text{otherwise}
        \end{cases} \right)
    \end{align*}
\end{lemma}
\begin{proof}
    First of all, we show that the map is well-defined. Let $f: (\Z/q\Z)^\units \to \C^\units$ be a (multiplicative) group homomorphism, and we show that $\chi_f$ is a Dirichlet character.

@@

>>>>

\title{Miniproject - Analytic Number Theory}
\author{Student Number: 1462479}
\date{19.12.2021}

@@

\begin{document}
\maketitle

\subsection*{Introduction}
In this Miniproject, we want to generalize the notion of the Riemannian Zeta function $\zeta(s)$ to investigate the frequency of primes in arithmetic progressions.
The central notion to do so is to consider Dirichlet characters $\chi: \N \to \C$.
We define this kind of map and study some of its properties.

After that, we introduce the Dirichlet series corresponding to a Dirichlet character, denoted by $L(s, \chi)$.
Using this series, we follow a path somewhat similar to what was done in the lecture to show an important part of Dirichlet's prime number theorem.

\subsection*{Conventions}
We use the convention that $\N = \{ n \in \Z \ | \ n \geq 0 \}$.
Further, we write $a \divides b$ if $a$ divides $b$ and $a \perp b$ if $a$ and $b$ are coprime.
Finally, let $\primes$ be the set of prime numbers in $\N$.

\section{Part I}
For convenience, we include the definition of a Dirichlet character from the task description first.
\begin{definition}
    Let $q \geq 2$, then a \emph{Dirichlet character (modulo $q$)} is a function $\chi: \N \to \C$ such that
    \begin{itemize}
        \item $\chi$ is completely multiplicative, so $\chi(a)\chi(b) = \chi(ab)$
        \item $\chi$ is periodic modulo $q$, so $\chi(n + q) = \chi(n)$
        \item $\chi(n) \neq 0$ if and only if $n \perp q$
    \end{itemize}
\end{definition}
First, we will give another characterization of Dirichlet characters.
\begin{lemma}[Characterization of Dirichlet characters]
    \label{prop:characterization_dirichlet_character}
    We have a one-to-one correspondence between Dirichlet characters modulo $q$ and group homomorphisms $(\Z/q\Z)^\units \to \C^\units$ via
    \begin{align*}
        \{ f: (\Z/q\Z)^\units \to \C^\units \ | \ f \ \text{group hom} \} &\to \{ \chi: \N \to \C \ | \ \chi \ \text{Dirichlet character mod $q$} \} \\
        f &\mapsto \chi_f := \left( \N \to \C, \ n \mapsto \begin{cases}
            f([n]) & \text{if $n \perp q$} \\
            0 & \text{otherwise}
        \end{cases} \right)
    \end{align*}
\end{lemma}
\begin{proof}
    First of all, we show that the map is well-defined. Let $f: (\Z/q\Z)^\units \to \C^\units$ be a (multiplicative) group homomorphism, and we show that $\chi_f$ is a Dirichlet character.

@@

\subsection*{Conclusion}
In this work, we have seen the basic structure of Dirichlet characters and their connection to the group $(\Z/q\Z)^\units$.
Furthermore, we have studied some properties of the series $L(s, \chi)$, in particular its analytic continuation to $\Re(s) > 0$ and an identiy for its logarithmic derivative.
By using the theory developed before, we could then establish a connection between the latter and the distribution of primes $\equiv a \pmod q$.
Assuming now a statement on the zeros of $L(s, \chi)$, we were able to prove Dirichlet's prime number theorem, namely that there are infinitely many primes $\equiv a \pmod q$ for all $a \perp q$ in the case of primes $q$.

Finally, we have also seen a very short detour into analyzing primes $\equiv 1 \mod q$ using Algebraic Number theory.
For this, we used the theory of cyclotomic number fields, which already hints at the strong connection of $L(s, \chi)$ to cyclotomic number fields that I could see in further literature.
Of course, also the encounter of the group $(\Z/q\Z)^\units$ and the connection of Dirichlet characters to roots of unity go into that direction.

Of course, the most central question when continuing on this path is to proof the fact $L(1, \chi) \neq 0$ for $\chi \neq \chi_0$ and thus complete the proof of Dirichlet's prime number theorem.
After that, it seems to be possible to use further techniques from the lecture (like Perron's formula) and show some kind of ``modular prime number theorem'', i.e. study the asymptotical frequency of primes $\equiv a \mod q$.
Since we already have a connection between $\Psi_a$ and $L(s, \chi)'/L(s, \chi)$ to which we could apply Perron's formula, we already have some ingredients for this proof.
To complete this, it would now be necessary to apply theorems like the ones from Chapter 9 in the lecture to $L(s, \chi)$, and then prove some kind of zero-free region for $L(s, \chi)$.
This would then show that
\begin{equation*}
    \rho_a(s) = \sum_{n \equiv a \ (\text{mod $q$})} \Lambda(n) n^{-s} = \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \frac {L(s, \chi)'} {L(s, \chi)}
\end{equation*}
has only one simple pole at $1$ in the zero-free region.
Using Cauchy's Residue theorem and moving the left contour of integration just left of the line $\Re(s) = 1$, we could then show a statement like $|\{ p \in \primes \ | \ p \equiv a \pmod q, p \leq x\}| \sim x / \log(x) / (q - 1)$, or whatever statement holds on the asymptotic frequency of primes in arithmetic progressions.

All in all, we have seen how to study primes in congruence classes modulo $q$ using analytic techniques.
As in the lecture, the next questions mainly deal with the distribution of zeros of $L(s, \chi)$.
